---
title: "PC4"
author: "Jack Walton, Calista Kerins, Dylan Schmidt, Jhon Chavez-Matul"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
    number-sections: true
    theme:
      light: flatly
      dark: darkly
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

# Project Proposal + Data

## **Quantitative Variables**

-   Variable 1: Happiness score (WHR)\*

-   Variable 2: Life Expectancy\*

\*Measured each year. Country is also recorded.\

## **Detailed Data Description**

[Happiness Score Dataset](https://docs.google.com/spreadsheets/d/1ll4dtbYCDKWzqIfpC5ZXYdxIyjZ8lKnEyWHQI4a_yM0/edit#gid=501532268): This data set contains average scores about life satisfaction using a scale known as the Cantril life ladder, measured across nations over multiple years. Respondents imagined a ladder with steps ranging from 0 at the bottom (representing the worst possible life) to 10 at the top (representing the best possible life) and indicated where they feel they stand. Gapminder has converted this scale from 0 to 100 for clarity, representing it in terms of percentage. The data comes from the World Happiness Report collected by the Gallup World Poll. It contains data from 2005-2020. You can also find out more [here](https://worldhappiness.report/ed/2023/#appendices-and-data) from the World Happiness Report, where this data originated from, since there is no documentation on Gapminder.

[Life Expectancy Dataset](https://www.gapminder.org/data/documentation/gd004/): This data set contains life expectancy across different nations over time, measured as the number of years a newborn infant would live if the current mortality rates at different ages were to stay the same throughout its life. The data is compiled by Gapminder and mainly comes from the Institute of Health Metrics and Evaluation and UN forecasts. It contains data from 1800 to 2100 (Predicted).

**Hypothesized Relationship**

```         
We hypothesize that a higher happiness score will also have a high life expectancy. We assume people who are happier try to stay healthy and take better care of themselves because they enjoy their lives.
```

## **Data Cleaning Process**


1.  First, we did an inner join on column ‘Year’ (same name in both datasets) to combine both tables, and keep only the rows where data was collected for both tables during that year.

2. Next, we renamed the variable for happiness score from `Happiness.score..WHR.` to `Happiness.score`

```{r}
library(tidyverse)
life_expectancies_by_country <- read.csv("Life Expectancy-Dataset-countries-etc-by-year.csv")
happiness_scores_by_country <- read.csv("Happiness-Dataset-countries-by-year.csv")

full_dataset <- inner_join(life_expectancies_by_country, 
                           happiness_scores_by_country, 
                           by = c("time", "name")) |>
                rename(Happiness.score = Happiness.score..WHR.)

```

# Linear Regression

Explanatory Variable (x): Happiness score

Response Variable (y): Life expectancy

## Data Visualization

Let's plot the data with the explanatory variable, happiness score, on the x-axis and the response variable, life expectancy, on the y-axis. We do this since it makes more sense that happiness would affect life expectancy rather than life expectancy affecting happiness.

```{r}
plot <- ggplot(data= full_dataset, aes(x= Happiness.score, y=Life.expectancy)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  labs(x = "Happiness score", y = "", subtitle = "Life Expectancy (Years)", title = "Relation Between Happiness and Life Expectancy")
plot(plot)
```

Excluding the outliers, the points seem to follow a somewhat linear trend. Now we'll separate the points by years and animate a gif of similar plots for each year.

```{r}
library(gganimate)
library(gifski)

anim <- plot +
  transition_time(time) + 
  labs(title = "Relation Between Happiness and Life Expectancy: {frame_time}")

animate(anim, renderer = gifski_renderer("happiness_life_expectancy.gif"))
```

The plots in the gif above all seem to keep the same trend line, so the linear relationship seems to be stable over time.

## Fitting a simple linear regression model to check for a relationship:

First, we cleaned the data to ensure one x value and one y value per country. We chose to use the average values for life expectancy and happiness scores for each country across **all** years, in order to get a more complete picture.

```{r}
clean_dataset <- full_dataset |>
  group_by(name) |>
  summarise(LifeExpectancy = mean(Life.expectancy), 
            HappinessScore = mean(Happiness.score))
  

model <- lm(LifeExpectancy ~ HappinessScore, clean_dataset)

model
```

$$
\text{Life Expectancy} = \beta_0 + \beta_1 \cdot \text{Happiness Score}
$$From the summary output we find that the y-intercept $\beta_0$ is $39.972$ and the slope $\beta_1$ is $0.586$. Thus, the estimated regression equation is: $$
\text{Life Expectancy} = 39.972 + 0.586 \cdot \text{Happiness Score}
$$ {#eq-Regression}

The equation seen above represents the linear regression line. This suggests a unit increase in happiness score corresponds to a 0.586 year increase in life expectancy.

## Model Fit

In order to asses the model fit, we determined the variance in the observed response values, the fitted values, and the residuals. These are presented in the table below.

```{r}
vars <- broom::augment(model) |>
  summarize(response_var = var(LifeExpectancy),
            fitted_var = var(.fitted),
            resid_var = var(.resid))

vars |>
  knitr::kable(col.names = c("Response Variance", "Fitted Variance", "Residuals Variance"),
               format = "pipe",
               align = "ccc",
               caption = "Table 2.3.1: Model Variances",
               digits = 2)
```

According to these variances, the model accounts for `r round((vars$fitted_var / vars$response_var) * 100, 2)`% of the variability in the response values. This suggests that our model does a moderately good job of representing the true response.
