---
title: "PC5: Final Report"
author: "Jack Walton, Calista Kerins, Dylan Schmidt, Jhon Chavez-Matul"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
    number-sections: true
    code-fold: true
    theme:
      light: flatly
      dark: darkly
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

# Project Proposal + Data

## Quantitative Variables

-   Variable 1: Happiness score (0-100)\*

-   Variable 2: Life Expectancy (Years; Recorded at birth)\*

\*Measured each year. Country is also recorded.\

## Detailed Data Description

[Happiness Score Dataset](https://docs.google.com/spreadsheets/d/1ll4dtbYCDKWzqIfpC5ZXYdxIyjZ8lKnEyWHQI4a_yM0/edit#gid=501532268): This data set contains average scores about life satisfaction using a scale known as the Cantril life ladder, measured across nations over multiple years. Respondents imagined a ladder with steps ranging from 0 at the bottom (representing the worst possible life) to 10 at the top (representing the best possible life) and indicated where they feel they stand. Gapminder has converted this scale from 0 to 100 for clarity, representing it in terms of percentage. The data comes from the World Happiness Report collected by the Gallup World Poll. It contains data from 2005-2020. You can also find out more [here](https://worldhappiness.report/ed/2023/#appendices-and-data) from the World Happiness Report, where this data originated from, since there is no documentation on Gapminder.

[Life Expectancy Dataset](https://www.gapminder.org/data/documentation/gd004/): This data set contains life expectancy across different nations over time, measured as the number of years a newborn infant would live if the current mortality rates at different ages were to stay the same throughout its life. The data is compiled by Gapminder and mainly comes from the Institute of Health Metrics and Evaluation and UN forecasts. It contains data from 1800 to 2100 (Predicted).

### Hypothesized Relationship

**The hypothesis posited is that a higher happiness score correlates with a higher life expectancy. The rationale is that individuals who are happier are more likely to stay healthy and take better care of themselves due to their enjoyment of life. Additionally, it is assumed that individuals born into poverty are less happy due to reduced access to health and wellness.**

## Data Cleaning Process

The data cleaning process commenced with an inner join on the column ‘Year’ (same name in both datasets) to merge both tables and retain only the rows where data was collected for both datasets during that year. Subsequently, the variable for happiness score was renamed from Happiness.score..WHR. to Happiness.score. There was no need for additional data cleaning steps such as pivoting longer or filling NA values as the datasets were already well-prepared.

```{r}
library(tidyverse)
life_expectancies_by_country <- read.csv("Life Expectancy-Dataset-countries-etc-by-year.csv")
happiness_scores_by_country <- read.csv("Happiness-Dataset-countries-by-year.csv")

full_dataset <- inner_join(life_expectancies_by_country, 
                           happiness_scores_by_country, 
                           by = c("time", "name")) |>
                rename(Happiness.score = Happiness.score..WHR.)

```

# Linear Regression

Explanatory Variable (x): Happiness score

Response Variable (y): Life expectancy

## Data Visualization

The relationship between the explanatory variable, happiness score, and the response variable, life expectancy, is visualized with the happiness score on the x-axis and life expectancy on the y-axis. This choice is based on the logical premise that happiness is more likely to affect life expectancy than vice versa.

```{r}
plot <- ggplot(data= full_dataset, aes(x= Happiness.score, y=Life.expectancy)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  labs(x = "Happiness score", y = "", subtitle = "Life Expectancy (Years)", title = "Relation Between Happiness and Life Expectancy")
plot(plot)
```

In the plot above, each point represents the happiness score and life expectancy from our full dataset. The trend suggests a positive relationship, where people with higher happiness scores tend to have higher life expectancies. This is an important visual confirmation of our hypothesized relationship. Excluding the outliers, the points seem to follow a somewhat linear trend, indicating that a linear regression might be the most appropriate model for this data.

Next, the further analysis below involves separating the points by years and animating a gif of similar plots for each year to observe any changes or trends over time.

```{r}
library(gganimate)
library(gifski)

anim <- plot +
  transition_time(time) + 
  labs(title = "Relation Between Happiness and Life Expectancy: {frame_time}")

animate(anim, renderer = gifski_renderer("happiness_life_expectancy.gif"))
```

The animated gif illustrates the relationship between happiness score and life expectancy over time. Each frame represents a different year, providing a dynamic view of whether the relationship holds consistently. The plots in the gif maintain a stable trend line over the years, reinforcing the stability of the linear relationship over time.

## Perform Linear Regression

### Average Values for Each Country

The data was summarized to ensure one happiness score (x value) and one life expectancy (y value) per country. The average values for life expectancy and happiness scores for each country across all years were used to provide a more comprehensive picture. This approach was chosen due to the stability of the linear model over time observed in the previous section.

``` {r}
clean_dataset <- full_dataset |>
  group_by(name) |>
  summarise(LifeExpectancy = mean(Life.expectancy), 
            HappinessScore = mean(Happiness.score))

avg_plot <- ggplot(data= clean_dataset, aes(x= HappinessScore, y=LifeExpectancy)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  labs(x = "Average Happiness Score", y = "Average Life Expectancy (Years)", title = "Average Happiness Score vs. Average Life Expectancy")

plot(avg_plot)
```

The plot above demonstrates the average happiness score versus the average life expectancy for each country. Each point signifies a country, with the x-axis displaying the average happiness score and the y-axis displaying the average life expectancy. The positive trend indicates that countries with higher average happiness scores also tend to have higher average life expectancies, further supporting the hypothesized relationship and providing a clearer, less cluttered visualization of the data. This visualization is crucial as it sets the stage for the subsequent linear regression model.

### Fitting a Simple Linear Regression Model:

A simple linear regression model was fitted to examine the relationship between happiness scores (independent/explanatory variable) and life expectancy (dependent/response variable):

```{r}
model <- lm(LifeExpectancy ~ HappinessScore, clean_dataset)

model
```

### Linear Regression Explanation
**Refresher:** Linear regression is a statistical method used to model the relationship between a dependent variable (response) and one or more independent variables (explanatories). In our case, the goal is to find a linear equation that best predicts the response variable (life expectancy) based on the explanatory variable (happiness score). 

The linear equation for our relationship is of the form:

$$
\text{Life Expectancy} = \beta_0 + \beta_1 \cdot \text{Happiness Score}
$$ {#eq-Regression}
From the summary output above, the y-intercept $\beta_0$ is $39.972$ and the slope $\beta_1$ is $0.586$. Thus, the estimated regression equation is: 
$$
\text{Life Expectancy} = 39.972 + 0.586 \cdot \text{Happiness Score}
$$ {#eq-Regression-Complete}

The completed linear equation seen above contains a slope which suggests a unit increase in happiness score corresponds to a 0.586 year increase in life expectancy. The y-intercept of the equation suggests that someone with a happiness score of 0 would have a life expectancy of approximately 40 years. 

## Model Fit

To assess the model fit, we determined the variance in the observed response values (life expectancy), the fitted values (predicted life expectancy from the model), and the residuals (the differences between observed and fitted values). These are presented in the table below.

```{r}
vars <- broom::augment(model) |>
  summarize(response_var = var(LifeExpectancy),
            fitted_var = var(.fitted),
            resid_var = var(.resid))

vars |>
  knitr::kable(col.names = c("Response Variance", "Fitted Variance", "Residuals Variance"),
               format = "pipe",
               align = "ccc",
               caption = "Table 2.3.1: Model Variances",
               digits = 2)
```

### Explanation of Variances:

**Response Variance:** This is the variance in the observed life expectancy values. It measures how much the life expectancy values vary around their mean.

**Fitted Variance:** This is the variance in the predicted life expectancy values from our linear regression model. It measures how much the predicted values vary around the mean of the observed values.

**Residuals Variance:** This is the variance in the residuals (the differences between the observed and predicted life expectancy values). It measures the amount of variation in the response variable that is not explained by the predictor variable in the model.

To determine the proportion of variance in life expectancy explained by the happiness score, the R-squared value is calculated by dividing the fitted variance by the response variance and multiplying by 100 to express it as a percentage:

$$
R^2 = \frac{\text{Fitted Variance}}{\text{Response Variance}} \times 100 = \frac{38.69}{62.13} \times 100 \approx 62.27\%
$$

According to these variances, the model accounts for `r round((vars$fitted_var / vars$response_var) * 100, 2)`% of the variability in the response values. This suggests that the linear regression model does a moderately good job of representing the true response, indicating a meaningful relationship between happiness scores and life expectancy. 

# Simulation

Performing predictive checks is essential to determine how well the observed data and the linear model fit. By comparing data generated from the linear model to the observed data, we can assess the plausibility that the observed data could be generated by the model.

To conduct a predictive check, the following steps are necessary:

  1. Fit a regression model to the observed data.

  2. For a set of observed explanatory values, obtain predicted response values from the model.

  3. Add random errors to the predicted responses.

  4. Compare the simulated data to the observed data.

  5. Iterate!
  
Step 1 has been completed in the previous section, Linear Regression. Next, predicted response values (life expectancy) will be obtained from the model using the observed explanatory values (happiness score).

## Visualizing Simulations from the Model
In simple linear regression, responses are modeled as a linear function of the explanatory variable and some error, represented by the equation:

$$

\text{Life Expectancy} = \beta_0 + \beta_1 \cdot \text{Happiness Score} + \varepsilon

$$ {#eq-Regression-with-err}

For simple linear regression, the errors ($\varepsilon$) are assumed to follow a normal distribution with a mean of 0 and a standard deviation.

First, predictions based on the linear regression model are generated. Then, observations are simulated by adding random errors to the predictions, based on the model's residual standard error and a mean of 0. With steps 2 and 3 of the predictive check completed, the observed values from the dataset are plotted next to the simulated response values for comparison below:

```{r}
set.seed(13)  # Purely for reproducibility

predicted_values <- predict(model, clean_dataset) # predict response values 

# Add random error based on residual std error of the model
residual_standard_error <- sigma(model)
simulated_values <- predicted_values + rnorm(length(predicted_values), mean = 0, sd = residual_standard_error)
clean_dataset$SimulatedLifeExpectancy <- simulated_values

# Combine observed and simulated data into one dataset
combined_dataset <- clean_dataset |>
  select(HappinessScore, LifeExpectancy) |>
  rename(Life_Expectancy = LifeExpectancy) |>
  mutate(Data_Type = "Observed") |>
  bind_rows(clean_dataset |>
              select(HappinessScore, SimulatedLifeExpectancy) |>
              rename(Life_Expectancy = SimulatedLifeExpectancy) |>
              mutate(Data_Type = "Simulated"))

# Plot using facet_wrap
combined_plot <- ggplot(data = combined_dataset, aes(x = HappinessScore, y = Life_Expectancy)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  labs(x = "Avg. Happiness Score", y = "", subtitle="Avg. Life Expectancy (Years)") +
  facet_wrap(~ Data_Type, scales = "free_y") +
  theme_minimal() +
  labs(title = "Comparison of Observed and Simulated Data: Happiness Score vs. Life Expectancy")

plot(combined_plot)
```
The plots above display the observed data (left) and the simulated data (right) side-by-side. The simulated data shows a similar positive trend as the observed data, indicating that the model's predictions are reasonably aligned with the actual observations. However, some differences can be observed:

**Variance:** The simulated data seems to exhibit a slightly thinner spread compared to the observed data, which has slightly denser grouping of points in certain areas of the plot. This is believed to be reflecting the random errors added to the predictions.
**Outliers:** Outliers present in the observed data appear more prominent than in the simulated data. For example, there are observed points with life expectancy under 50 years.

## Generating Multiple Predictive Checks
To rigorously test the model, generating at least 1,000 simulated datasets is essential. These datasets could have been observed if the regression model adequately describes the relationship between the response and explanatory variables.

In step 4, a linear regression is performed between the observed dataset and each simulated dataset, and the $R^2$ value from each regression is recorded. This process is iterated 1,000 times in total. Understanding how well the model can generate data similar to the observed dataset is crucial. The distribution of these $R^2$ values is then plotted with their frequencies.

```{r}
library(purrr)

set.seed(13)  # For reproducibility
num_simulations <- 1000

# Function to simulate and calculate R-squared
simulate_r2 <- function(predicted_values, residual_standard_error, observed_values) {
  simulated_values <- predicted_values + rnorm(length(predicted_values), mean = 0, sd = residual_standard_error)
  sim_model <- lm(observed_values ~ simulated_values)
  summary(sim_model)$r.squared
}

# Generate R-squared values
simulated_r2 <- map_dbl(1:num_simulations, ~simulate_r2(predicted_values, residual_standard_error, clean_dataset$LifeExpectancy))

# Create data frame for plotting
simulated_r2_df <- data.frame(R2 = simulated_r2)

# Plot distribution of R-squared values
r2_plot <- ggplot(simulated_r2_df, aes(x = R2)) +
  geom_histogram(binwidth = 0.01, fill = "dodgerblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of R² from Simulated Data", x = "R²", y = "Frequency")

plot(r2_plot)
```
The histogram above illustrates the distribution of $R^2$ values obtained from 1,000 simulated datasets. The key observation here is the clustering of the majority of $R^2$ values around a central value of 0.4, indicating a consistent level of model performance across simulations.

### Implications of an Average $R^2$ Value of 0.4
#### Predictive Power
An average $R^2$ value of 0.4 indicates that the model has a moderate level of predictive power. It explains a significant portion, but not the majority, of the variation in life expectancy. This suggests that while happiness score is an important predictor of life expectancy, other factors contribute to life expectancy that are not captured by this linear regression model.

#### Model Performance and Variance
The $R^2$ value reflects the model's ability to fit the data. With 40% of the variance explained, the model performs reasonably well, but there is room for improvement. This moderate fit indicates that the linear relationship between happiness score and life expectancy is meaningful, but the model does not capture all the complexity of the data. The remaining 60% of the variance in life expectancy is unexplained by the model.

### Conclusion
While an average $ R^2 $ value of 0.4 indicates that happiness score is a significant and meaningful predictor of life expectancy, it also highlights the complexity of the factors influencing life expectancy. For example, factors such as socioeconomic status, healthcare access, education, and environmental conditions might also play significant roles. The model's moderate predictive power encourages further research to identify additional variables and refine the model for a more comprehensive understanding of the determinants of life expectancy. Exploring non-linear models also may provide a better fit and more insights into the relationship between happiness and life expectancy.

