---
title: "PC5: Final Report"
author: "Jack Walton, Calista Kerins, Dylan Schmidt, Jhon Chavez-Matul"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
    number-sections: true
    code-fold: true
    theme:
      light: flatly
      dark: darkly
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

# Project Proposal + Data

## Quantitative Variables

-   Variable 1: Happiness score (0-100)\*

-   Variable 2: Life Expectancy (Years; Recorded at birth)\*

\*Measured each year. Country is also recorded.\

## Detailed Data Description

[Happiness Score Dataset](https://docs.google.com/spreadsheets/d/1ll4dtbYCDKWzqIfpC5ZXYdxIyjZ8lKnEyWHQI4a_yM0/edit#gid=501532268): This data set contains average scores about life satisfaction using a scale known as the Cantril life ladder, measured across nations over multiple years. Respondents imagined a ladder with steps ranging from 0 at the bottom (representing the worst possible life) to 10 at the top (representing the best possible life) and indicated where they feel they stand. Gapminder has converted this scale from 0 to 100 for clarity, representing it in terms of percentage. The data comes from the World Happiness Report collected by the Gallup World Poll. It contains data from 2005-2020. You can also find out more [here](https://worldhappiness.report/ed/2023/#appendices-and-data) from the World Happiness Report, where this data originated from, since there is no documentation on Gapminder.

[Life Expectancy Dataset](https://www.gapminder.org/data/documentation/gd004/): This data set contains life expectancy across different nations over time, measured as the number of years a newborn infant would live if the current mortality rates at different ages were to stay the same throughout its life. The data is compiled by Gapminder and mainly comes from the Institute of Health Metrics and Evaluation and UN forecasts. It contains data from 1800 to 2100 (Predicted).

### Hypothesized Relationship

**We hypothesize that a higher happiness score will also have a high life expectancy. We assume people who are happier try to stay healthy and take better care of themselves because they enjoy their lives. We also assume that people born into poverty are less happy, and have less access to health and wellness.**

## Data Cleaning Process


1.  First, we did an inner join on column ‘Year’ (same name in both datasets) to combine both tables, and keep only the rows where data was collected for both tables during that year.

2. Next, we renamed the variable for happiness score from `Happiness.score..WHR.` to `Happiness.score`

3. Note: We did not need to pivot_longer or fill NA values because our datasets were as clean as possible to begin with.

```{r}
library(tidyverse)
life_expectancies_by_country <- read.csv("Life Expectancy-Dataset-countries-etc-by-year.csv")
happiness_scores_by_country <- read.csv("Happiness-Dataset-countries-by-year.csv")

full_dataset <- inner_join(life_expectancies_by_country, 
                           happiness_scores_by_country, 
                           by = c("time", "name")) |>
                rename(Happiness.score = Happiness.score..WHR.)

```

# Linear Regression

Explanatory Variable (x): Happiness score

Response Variable (y): Life expectancy

## Data Visualization

Let's plot the data with the explanatory variable, happiness score, on the x-axis and the response variable, life expectancy, on the y-axis. We do this since it makes more sense that happiness would affect life expectancy rather than life expectancy affecting happiness.

```{r}
plot <- ggplot(data= full_dataset, aes(x= Happiness.score, y=Life.expectancy)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  labs(x = "Happiness score", y = "", subtitle = "Life Expectancy (Years)", title = "Relation Between Happiness and Life Expectancy")
plot(plot)
```

In the plot above, each point represents the happiness score and life expectancy from our full dataset. The trend suggests a positive relationship, where people with higher happiness scores tend to have higher life expectancies. This is an important visual confirmation of our hypothesized relationship. Excluding the outliers, the points seem to follow a somewhat linear trend, indicating that a linear regression might be the most appropriate model for this data.

Now we'll separate the points by years and animate a gif of similar plots for each year. This will help us observe any changes or trends over time.

```{r}
library(gganimate)
library(gifski)

anim <- plot +
  transition_time(time) + 
  labs(title = "Relation Between Happiness and Life Expectancy: {frame_time}")

animate(anim, renderer = gifski_renderer("happiness_life_expectancy.gif"))
```

The animated gif created above shows the relationship between happiness score and life expectancy over time. Each frame represents a different year, allowing us to see if the relationship holds consistently. The plots in the gif seem to keep the same trend line over the years, suggesting that the linear relationship is stable over time.

## Perform Linear Regression

### Average Values for Each Country

First, we summarized the data to ensure one happiness score (x value) and one life expectancy (y value) per country. We chose to use the average values for life expectancy and happiness scores for each country across **all** years, in order to get a more complete picture. We decided to average across all years because the linear model stayed stable over time in the previous section.

``` {r}
clean_dataset <- full_dataset |>
  group_by(name) |>
  summarise(LifeExpectancy = mean(Life.expectancy), 
            HappinessScore = mean(Happiness.score))

avg_plot <- ggplot(data= clean_dataset, aes(x= HappinessScore, y=LifeExpectancy)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  labs(x = "Average Happiness Score", y = "Average Life Expectancy (Years)", title = "Average Happiness Score vs. Average Life Expectancy")

plot(avg_plot)
```

The plot above shows the average happiness score versus the average life expectancy for each country. Each point represents a country, with the x-axis displaying the average happiness score and the y-axis displaying the average life expectancy. The positive trend observed in the plot indicates that countries with higher average happiness scores also tend to have higher average life expectancies. This further supports our hypothesized relationship and provides a clearer, less cluttered visualization of the data.
It's also important to visualize this data, because next we will use for our linear regression model.

### Fitting a Simple Linear Regression Model:

Now let's fit a simple linear regression model to check for a relationship between happiness scores (dependent/response variable) and life expectancy (independent/explanatory variable):

```{r}
model <- lm(LifeExpectancy ~ HappinessScore, clean_dataset)

model
```

### Linear Regression Explanation
**Refresher:** Linear regression is a statistical method used to model the relationship between a dependent variable (response) and one or more independent variables (explanatories). In our case, the goal is to find a linear equation that best predicts the response variable (life expectancy) based on the explanatory variable (happiness score). 

The linear equation for our relationship is of the form:

$$
\text{Life Expectancy} = \beta_0 + \beta_1 \cdot \text{Happiness Score}
$$ {#eq-Regression}
From the summary output above, we find that the y-intercept $\beta_0$ is $39.972$ and the slope $\beta_1$ is $0.586$. Thus, the estimated regression equation is: 
$$
\text{Life Expectancy} = 39.972 + 0.586 \cdot \text{Happiness Score}
$$ {#eq-Regression-Complete}

The completed linear equation seen above contains a slope which suggests a unit increase in happiness score corresponds to a 0.586 year increase in life expectancy. The y-intercept of the equation suggests that someone with a happiness score of 0 would have a life expectancy of ~40 years. 

## Model Fit

To assess the model fit, we determined the variance in the observed response values (life expectancy), the fitted values (predicted life expectancy from the model), and the residuals (the differences between observed and fitted values). These are presented in the table below.

```{r}
vars <- broom::augment(model) |>
  summarize(response_var = var(LifeExpectancy),
            fitted_var = var(.fitted),
            resid_var = var(.resid))

vars |>
  knitr::kable(col.names = c("Response Variance", "Fitted Variance", "Residuals Variance"),
               format = "pipe",
               align = "ccc",
               caption = "Table 2.3.1: Model Variances",
               digits = 2)
```

### Explanation of Variances:

**Response Variance:** This is the variance in the observed life expectancy values. It measures how much the life expectancy values vary around their mean.

**Fitted Variance:** This is the variance in the predicted life expectancy values from our linear regression model. It measures how much the predicted values vary around the mean of the observed values.

**Residuals Variance:** This is the variance in the residuals (the differences between the observed and predicted life expectancy values). It measures the amount of variation in the response variable that is not explained by the predictor variable in the model.

To determine the proportion of variance in life expectancy explained by the happiness score, we calculate the R-squared value. This is done by dividing the fitted variance by the response variance and then multiplying by 100 to express it as a percentage:

$$
R^2 = \frac{\text{Fitted Variance}}{\text{Response Variance}} \times 100 = \frac{38.69}{62.13} \times 100 \approx 62.27\%
$$

According to these variances, the model accounts for `r round((vars$fitted_var / vars$response_var) * 100, 2)`% of the variability in the response values. This suggests that our model does a moderately good job of representing the true response.

# Simulation

## Visualizing Simulations from the Model

## Generating Multiple Predictive Checks
